Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference

### 2017-11-29 workflow/workflow.Rmd
add SingleCellExperiment

```{r options, echo=FALSE, results="hide",message=FALSE, error=FALSE, include=FALSE, autodep=TRUE}
if(packageVersion("zinbwave")<"0.99.6") stop("must have current devel version to avoid bugs")
if(packageVersion("clusterExperiment")<"1.3.2") stop("must have current devel version to avoid bugs")
if(packageVersion("scone")<"1.1.2") stop("must have current devel version to avoid bugs")

@@

se <- SummarizedExperiment(assays = list(counts = E),
                           colData = metadata)
se
```

@@

```{r scone, fig.cap="SCONE: Filtering of low-quality cells."}
# QC-metric-based sample-filtering
data("housekeeping")
hk = rownames(se)[toupper(rownames(se)) %in% housekeeping$V1]

mfilt <- metric_sample_filter(assay(se), 
                              nreads = colData(se)$NREADS,
                              ralign = colData(se)$RALIGN,
                              pos_controls = rownames(se) %in% hk,
                              zcut = 3, mixture = FALSE,
                              plot = TRUE)
```

```{r sconeFilt}
# Simplify to a single logical
mfilt <- !apply(simplify2array(mfilt[!is.na(mfilt)]), 1, any)
se <- se[, mfilt]
dim(se)
```

After sample filtering, we are left with `r ncol(se)` good quality cells.

@@

```{r}
# Filtering to top 1,000 most variable genes
vars <- rowVars(log1p(assay(se)))
names(vars) <- rownames(se)
vars <- sort(vars, decreasing = TRUE)
core <- se[names(vars)[1:1000],]
```

@@

```{r zinb}
print(system.time(se <- zinbwave(core, K = 50, X = "~ Batch",  
                                 residuals = TRUE,
                                 normalizedValues = TRUE)))
```

@@

```{r norm}
norm <- assays(se)$normalizedValues
norm[1:3,1:3]
```

@@

@@

```{r rsec_50}
seObj <- SummarizedExperiment(t(W), colData = colData(core))
print(system.time(ceObj <- RSEC(seObj, k0s = 4:15, alphas = c(0.1),
                                betas = 0.8, dimReduce="none",
                                clusterFunction = "hierarchical01", minSizes=1,
                                ncores = NCORES, isCount=FALSE,
                                dendroReduce="none",dendroNDims=NA,
                                subsampleArgs = list(resamp.num=100,
                                                     clusterFunction="kmeans",
                                                     clusterArgs=list(nstart=10)),
                                verbose=TRUE,
                                combineProportion = 0.7,
                                mergeMethod = "none", random.seed=424242,
                                combineMinSize = 10)))
```

@@

```{r examineCombineMany, fig.cap="RSEC: Candidate clusterings found using the function RSEC from the clusterExperiment package."}
plotClusters(ceObj, colPalette = c(bigPalette, rainbow(199)))
```

@@

```{r plotcoclust, fig.cap="RSEC: Heatmap of co-clustering matrix."}
plotCoClustering(ceObj)
```

@@

```{r tableclust}
table(primaryClusterNamed(ceObj))
```

```{r barplotOurs, fig.cap="RSEC: Barplot of number of cells per cluster for our workflow's RSEC clustering."}
plotBarplot(ceObj, legend = FALSE) 
```

@@

```{r addPublishedClusters, fig.cap="RSEC: Barplot of number of cells per cluster, for our workflow's RSEC clustering, color-coded by original published clustering."}
ceObj <- addClusters(ceObj, colData(ceObj)$publishedClusters, 
                     clusterLabel = "publishedClusters")

## change default color to match with Figure 7
clusterLegend(ceObj)$publishedClusters[, "color"] <- 
  col_clus[clusterLegend(ceObj)$publishedClusters[, "name"]]

plotBarplot(ceObj, whichClusters=c("combineMany","publishedClusters"),
            xlab = "", legend = FALSE)
```

@@

```{r heatmapsClusters, fig.cap="RSEC: Heatmap of the normalized expression measures for the 1,000 most variable genes, where rows correspond to genes and columns to cells ordered by RSEC clusters."}
# Set colors for cell clusterings
colData(ceObj)$publishedClusters <- as.factor(colData(ceObj)$publishedClusters)
origClusterColors <- bigPalette[1:nlevels(colData(ceObj)$publishedClusters)]
experimentColors <- bigPalette[1:nlevels(colData(ceObj)$Experiment)]
batchColors <- bigPalette[1:nlevels(colData(ceObj)$Batch)]
metaColors <- list("Experiment" = experimentColors,
                   "Batch" = batchColors,
                   "publishedClusters" = origClusterColors)

plotHeatmap(ceObj, visualizeData = assays(se)$normalizedValues,
            whichClusters = "primary", clusterFeaturesData = "all",
            clusterSamplesData = "dendrogramValue", breaks = 0.99,
            sampleData = c("publishedClusters", "Batch", "Experiment"),
            clusterLegend = metaColors, annLegend = FALSE, main = "")
```

@@

```{r mdsWce, fig.cap="RSEC: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by RSEC clustering."}
palDF <- ceObj@clusterLegend[[1]]
pal <- palDF[, "color"]
names(pal) <- palDF[, "name"]
pal["-1"] = "transparent"
plot(fit$points, col = pal[primaryClusterNamed(ceObj)], main = "", pch = 20,
     xlab = "Component1", ylab = "Component2")
legend(x = "topleft", legend = names(pal), cex = .5,
       fill = pal, title = "Sample")
```

@@

```{r tabagain}
table(data.frame(original = publishedClusters, ours = primaryClusterNamed(ceObj)))
```

@@

```{r boxplotReg3g, fig.width=6, fig.height=4, fig.cap="RSEC: Boxplots of the log count of gene Reg3g stratified by cluster."}
c4 <- rep("other clusters", ncol(se)) 
c4[primaryClusterNamed(ceObj) == "c4"] <- "cluster c4"
boxplot(log1p(assay(se)["Reg3g", ]) ~ primaryClusterNamed(ceObj), 
        ylab = "Reg3g log counts", cex.axis = .8, cex.lab = .8)
```

@@

```{r}
our_cl <- primaryClusterNamed(ceObj)
cl <- our_cl[!our_cl %in% c("-1", "c4")]
pal <- pal[!names(pal) %in% c("-1", "c4")]
X <- W[!our_cl %in% c("-1", "c4"), ]
mds <- cmdscale(dist(X), eig = TRUE, k = 4)
X <- mds$points

lineages <- getLineages(X, clusterLabels = cl, start.clus = "c1")
```

@@

```{r tree, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space, with connecting lines between cluster centers representing the inferred global lineage structure."}
pairs(lineages, type="lineages", col = pal[cl])
```

@@

```{r curves, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space, with smooth curves representing each inferred lineage."}
lineages <- getCurves(lineages)
pairs(lineages, type="curves", col = pal[cl])
```

@@

```{r slingshotsupervised, eval=FALSE}
lineages <- getLineages(X, clusterLabels = cl, start.clus = "c1",
                         end.clus = c("c3", "c7"))
lineagees <- getCurves(lineages)
pairs(lineages, type="curves", col = pal[primaryClusterNamed(ceObj)])
pairs(lineages, type="lineages", col = pal[primaryClusterNamed(ceObj)], 
      show.constraints = TRUE)

@@

```{r fitgam}
t <- pseudotime(lineages)[,1]
y <- assays(se)$normalizedValues[, !our_cl %in% c("-1", "c4")]
gam.pval <- apply(y,1,function(z){
  d <- data.frame(z=z, t=t)
  tmp <- gam(z ~ lo(t), data=d)
  p <- summary(tmp)[4][[1]][1,5]
  p
})
```

```{r heatmapsignificant, fig.cap="DE: Heatmap of the normalized expression measures for the 100 most significantly DE genes for the neuronal lineage, where rows correspond to genes and columns to cells ordered by pseudotime."}
topgenes <- names(sort(gam.pval, decreasing = FALSE))[1:100]
heatdata <- y[rownames(se) %in% topgenes, order(t, na.last = NA)]
heatclus <- cl[order(t, na.last = NA)]
ce <- clusterExperiment(heatdata, heatclus, transformation = identity)

#match to existing colors
cols <- clusterLegend(ceObj)$combineMany[, "color"]
names(cols) <- clusterLegend(ceObj)$combineMany[, "name"]
clusterLegend(ce)$cluster1[, "color"] <- cols[clusterLegend(ce)$cluster1[, "name"]]

plotHeatmap(ce, clusterSamplesData = "orderSamplesValue", breaks = .99)
```

>>>>

```{r options, echo=FALSE, results="hide",message=FALSE, error=FALSE, include=FALSE, autodep=TRUE}
if(packageVersion("zinbwave")<"0.99.6") stop("must have current devel version to avoid bugs")
if(packageVersion("clusterExperiment")<"1.5.0-9005") stop("must have current devel version to avoid bugs")
if(packageVersion("scone")<"1.1.2") stop("must have current devel version to avoid bugs")

@@

dataObj <- SummarizedExperiment(assays = list(counts = E),
                           colData = metadata)
dataObj
```

EP: changing 'se' to 'dataObj' because it changes classes. Also can search and replace 'dataObj' and not 'se'

@@

```{r scone, fig.cap="SCONE: Filtering of low-quality cells."}
# QC-metric-based sample-filtering
data("housekeeping")
hk = rownames(dataObj)[toupper(rownames(dataObj)) %in% housekeeping$V1]

mfilt <- metric_sample_filter(assay(dataObj), 
                              nreads = colData(dataObj)$NREADS,
                              ralign = colData(dataObj)$RALIGN,
                              pos_controls = rownames(dataObj) %in% hk,
                              zcut = 3, mixture = FALSE,
                              plot = TRUE)
```

```{r sconeFilt}
# Simplify to a single logical
mfilt <- !apply(simplify2array(mfilt[!is.na(mfilt)]), 1, any)
dataObj <- dataObj[, mfilt]
dim(dataObj)
```

After sample filtering, we are left with `r ncol(dataObj)` good quality cells.

@@

```{r}
# Filtering to top 1,000 most variable genes
vars <- rowVars(log1p(assay(dataObj)))
names(vars) <- rownames(dataObj)
vars <- sort(vars, decreasing = TRUE)
core <- dataObj[names(vars)[1:1000],]
```

EP: If use SingleCellFilter (or davide adds to singleCellExperiment) could do this in 1 line.

EP: Why do we create 'core' rather than just replace 'dataObj'

@@

```{r zinb}
print(system.time(dataObj <- zinbwave(core, K = 50, X = "~ Batch",  
                                 residuals = TRUE,
                                 normalizedValues = TRUE)))
```

@@

```{r norm}
norm <- assays(dataObj)$normalizedValues
norm[1:3,1:3]
```

@@

We are going to set these normalized values to be our first assay, so that they become the default in future plots.

```{r setNormAssay}
assay(dataObj)<-assays(dataObj)$normalizedValues
```

EP: This saves a lot of trouble with the plotHeatmap, etc. The code needs to be fixed so *reorders* the assays, rather than what I've done, which is just replace the first. I don't have the object in front of me to be able to do it (it might make sense for me to add option to different assays in clusterExperiment, but I'm not going to do that right now.)

@@

```{r addW}
reducedDim(dataObj,"zinbwave50")<-W
```

EP: If zinbwave already has been updated, this above code is probably not needed since its aready there.

EP: Note that we need to see whether clusterExperiment retains all of the different assays or not. It should in principle, but its not something I've tested.

```{r rsec_50}
print(system.time(dataObj <- RSEC(dataObj, k0s = 4:15, alphas = c(0.1),
                                betas = 0.8, dimReduce="zinbwave50",
                                clusterFunction = "hierarchical01", minSizes=1,
                                ncores = NCORES, isCount=FALSE,
                                dendroReduce="none",dendroNDims=NA,
                                subsampleArgs = list(resamp.num=100,
                                                     clusterFunction="kmeans",
                                                     clusterArgs=list(nstart=10)),
                                verbose=TRUE,
                                combineProportion = 0.7,
                                mergeMethod = "none", random.seed=424242,
                                combineMinSize = 10)))
```

@@

```{r examineCombineMany, fig.cap="RSEC: Candidate clusterings found using the function RSEC from the clusterExperiment package."}
plotClusters(dataObj)
```

EP: I've updated so that the default has >500 colors, so don't need the colPalette command, I hope.

@@

```{r plotcoclust, fig.cap="RSEC: Heatmap of co-clustering matrix."}
plotCoClustering(dataObj)
```

@@

```{r tableclust}
table(primaryClusterNamed(dataObj))
```

```{r barplotOurs, fig.cap="RSEC: Barplot of number of cells per cluster for our workflow's RSEC clustering."}
plotBarplot(dataObj, legend = FALSE) 
```

@@

```{r addPublishedClusters, fig.cap="RSEC: Barplot of number of cells per cluster, for our workflow's RSEC clustering, color-coded by original published clustering."}
dataObj <- addClusters(dataObj, colData(dataObj)$publishedClusters, 
                     clusterLabel = "publishedClusters")

## change default color to match with Figure 7
clusterLegend(dataObj)$publishedClusters[, "color"] <- 
  col_clus[clusterLegend(dataObj)$publishedClusters[, "name"]]

plotBarplot(dataObj, whichClusters=c("combineMany","publishedClusters"),
            xlab = "", legend = FALSE)
```

@@

```{r heatmapsClusters, fig.cap="RSEC: Heatmap of the normalized expression measures for the 1,000 most variable genes, where rows correspond to genes and columns to cells ordered by RSEC clusters."}
# Set colors for cell clusterings
experimentColors <- bigPalette[1:nlevels(colData(dataObj)$Experiment)]
batchColors <- bigPalette[1:nlevels(colData(dataObj)$Batch)]
metaColors <- list("Experiment" = experimentColors,
                   "Batch" = batchColors)

plotHeatmap(dataObj, 
            whichClusters = "primary", clusterFeaturesData = "all",
            clusterSamplesData = "dendrogramValue", breaks = 0.99,
            sampleData = c("Batch", "Experiment"),
            clusterLegend = metaColors, annLegend = FALSE, main = "")
```

EP: publishedClusters is saved as a cluster, so don't need to set it up. Now normalized should be default assay, so not need 'visualizeData'

@@

```{r mdsWce, fig.cap="RSEC: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by RSEC clustering."}
plotDimReduce(dataObj,whichCluster="primary",dimReduce="zinbwave50",pch=20,
	xlab = "Component1", ylab = "Component2",legendTitle="Sample",main=""
)
```

@@

```{r tabagain}
table(data.frame(original = publishedClusters, ours = primaryClusterNamed(dataObj)))
```

@@

```{r boxplotReg3g, fig.width=6, fig.height=4, fig.cap="RSEC: Boxplots of the log count of gene Reg3g stratified by cluster."}
# c4 <- rep("other clusters", ncol(se))
# c4[primaryClusterNamed(dataObj) == "c4"] <- "cluster c4"
boxplot(log1p(assay(se)["Reg3g", ]) ~ primaryClusterNamed(dataObj), 
        ylab = "Reg3g log counts", cex.axis = .8, cex.lab = .8)
```

EP: not sure why needed 'c4' object. This is a useful plot and rather messy, so I might make a plotGeneBoxplot function.

@@

```{r}
pseudoCe<-dataObj[,!primaryClusterNamed(dataObj) %in% c("-1", "c4")]
X <- reducedDim(pseudoCe,type="zinbwave50")
mds <- cmdscale(dist(X), eig = TRUE, k = 4)
X <- mds$points
lineages <- getLineages(X, clusterLabels = primaryClusterNamed(pseudoCe), start.clus = "c1")
```

EP: created a pseudoCe to subset to the samples for the lineage. Simplifies code enormously and as long as we don't change it, doesn't use additional memory, I think.

@@

```{r tree, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space, with connecting lines between cluster centers representing the inferred global lineage structure."}
colorCl<-convertClusterLegend(pseudoCe,whichCluster="primary",output="matrixColors")
pairs(lineages, type="lineages", col = colorCl)
```

EP: convertClusterLegend does the transformation into vector of colors. However, output may be matrix, in which case need to add a `[,1]` to the above code.

@@

```{r curves, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space, with smooth curves representing each inferred lineage."}
lineages <- getCurves(lineages)
pairs(lineages, type="curves", col = colorCl)
```

@@

```{r slingshotsupervised, eval=FALSE}
lineages <- getLineages(X, clusterLabels = primaryClusterNamed(pseudoCe), start.clus = "c1",
                         end.clus = c("c3", "c7"))
lineagees <- getCurves(lineages)
pairs(lineages, type="curves", col = colorCl)
pairs(lineages, type="lineages", col = colorCl, 
      show.constraints = TRUE)

@@

```{r fitgam}
t <- pseudotime(lineages)[,1]
y <- assay(pseudoCe)
gam.pval <- apply(y,1,function(z){
  d <- data.frame(z=z, t=t)
  tmp <- gam(z ~ lo(t), data=d)
  p <- summary(tmp)[4][[1]][1,5]
  p
})
```

EP: Since changed NormalizedValues to the main assay, this cleans up the code above and below.

```{r heatmapsignificant, fig.cap="DE: Heatmap of the normalized expression measures for the 100 most significantly DE genes for the neuronal lineage, where rows correspond to genes and columns to cells ordered by pseudotime."}
topgenes <- names(sort(gam.pval, decreasing = FALSE))[1:100]
orderSamples(pseudoCe)<-order(t,na.last=TRUE)
plotHeatmap(pseudoCe[,!is.na(t)], clusterSamplesData = "orderSamplesValue", breaks = .99)
```

The above code should take care of the ordering and NA values. Note that subseting a ClusterExperiment object is suppose to also subset the orderSamples value correctly. Let me know if that is not the case.
